Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-04-14 paper/graph.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@546 3fc257bb-8550-0410-81b3-a5c103da83f7

In the map() operation, each of $P$ processors is tasked with
generating a $1/P$ fraction of the desired edges.  A single random
$i,j$ edge is computed recursively as follows.  Pick a random quadrant
of the $A$ matrix with relative probabilities $a$, $b$, $c$, and $d$.
Treat the chosen quadrant as a sub-matrix and select a random quadrant
within it, in the same manner.  Repeat this process $n$ times where $N
= 2^n$.  At the end of the recursion, the final ``quadrant'' is
non-zero matrix element $A_{ij}$.

@@

Note that the degree distribution of vertices in the graph depends on
the choice of parameters $a$, $b$, $c$, $d$.  If one of the four
values is larger than the other 3, a highly skewed distribution
results.  Variants of the above algorithm can be used when $N$ is not
a power-of-two, to generate graphs with weighted edges (assign a
random numeric value to the $A_{ij}$ edge), graphs without self edges
(require $i != j$), or graphs with undirected edges (require $i < $j).

@@

A triangle in a graph is any triplet of vertices $(V_i,V_j,V_k)$ where
the three edges $(V_i,V_j), (V_j,V_k), (V_i,V_k)$ exist.  Figure
\ref{fig:tri} outlines a MapReduce algorithm that enumerates all such
triplets, assuming an input graph of undirected edge keys $(V_i,V_j)$
where $V_i < V_j$ for every edge, i.e. an upper-triangular R-MAT
generated matrix.  This exposition follows the triangle-finding
algorithm presented in \cite{Cohen}.

@@

Discuss connected component identification algorithm.

Do we need differnt variants of CC?

Could just do one, discuss possible inefficiency due to giant CC.

@@

 \caption{MapReduce algorithm for connected component identification.}

@@

>>>>

In the map() operation, each of $P$ processors generates a $1/P$
fraction of the desired edges.  A single random $i,j$ edge is computed
recursively as follows.  Pick a random quadrant of the $A$ matrix with
relative probabilities $a$, $b$, $c$, and $d$.  Treat the chosen
quadrant as a sub-matrix and select a random quadrant within it, in
the same manner.  Repeat this process $n$ times where $N = 2^n$.  At
the end of the recursion, the final ``quadrant'' is non-zero matrix
element $A_{ij}$.

@@

Note that the degree distribution of vertices in the graph depends on
the choice of parameters $a$, $b$, $c$, $d$.  If one of the four
values is larger than the other 3, a highly skewed distribution
results.  Variants of the above algorithm can be used when $N$ is not
a power-of-two, to generate graphs with weighted edges (assign a
numeric value to the $A_{ij}$ edge), graphs without self edges
(require $i != j$), or graphs with undirected edges (require $i < $j).
Or the general R-MAT matrix can be further processed by MapReduce
operations to meet these requirements.

@@

A triangle in a graph is any triplet of vertices $(V_i,V_j,V_k)$ where
the edges $(V_i,V_j), (V_j,V_k), (V_i,V_k)$ exist.  Figure
\ref{fig:tri} outlines a MapReduce algorithm that enumerates all
triangles, assuming an input graph of undirected edges $(V_i,V_j)$
where $V_i < V_j$ for every edge, i.e. an upper-triangular R-MAT
matrix.  This exposition follows the triangle-finding algorithm
presented in \cite{Cohen}.

@@

A connected component of a graph is a set of vertices where all pairs
of vertices in the set are connected by a path of edges.  A sparse
graph may contain many such components.  Figure \ref{fig:cc} outlines
a MapReduce algorithm that labels each vertex in a graph with a
component ID.  All vertices in the same component are labelled with
the same ID, which is the ID of a vertex in the component.  We assume
an input graph of undirected edges $(V_i,V_j)$.  This exposition also
follows the connected-component algorithm presented in \cite{Cohen},
with the addition of logic that load-balances data across processors
when one or a few giant components exist in the graph.

\begin{figure}[htb]
 \begin{center}\fbox{\begin{minipage}{\textwidth} \begin{tabbing}
  xxxx\=xxxxxxxxxxx\=xxxx\=xxxx\=xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx\= \kill

Iterate: \\
\> 1 Map: \> Convert edges to vertices \\
    \> \> \> input Key = $E_{ij}$, Value = NULL \\
    \> \> \> output Key = $V_i$, Value = $E_{ij}$ \\
    \> \> \> output Key = $V_j$, Value = $E_{ij}$ \\
\> 1 Add: \> Zone assignment of each vertex \\
    \> \> \> output Key = $V_i$, Value = $Z_i$ \\
\> 1 Collate: \> Vertex as key \\
\> 1 Reduce: \> Emit edges of each vertex with zone of vertex \\
       \> \> \> input Key = $V_i$, MultiValue = $E E E E ... Z$ \\
       \> \> \> for each $E$ in MultiValue: \\
      \> \> \> \> output Key = $E_{ij}$, Value = $Z_i$ \\

\> 2 Collate: \> Edge as key \\
\> 2 Reduce: \> Emit zone re-assignments \\
       \> \> \> input Key = $E_{ij}$, MultiValue = $Z_i Z_j$ \\
       \> \> \> $Z_{winner}$ = min($Z_i$,$Z_j$); $Z_{loser}$ = max($Z_i$,$Z_j$) \\
       \> \> \> if $Z_i$ and $Z_j$ are different: \\
      \> \> \> \> output Key = $Z_{loser}$, Value = $Z_{winner}$ \\
\> 2 Exit: \> if no output by Reduce 2 \\
\> 3 Map: \> Invert vertex/zone pairs \\
    \> \> \> input Key = $V_i$, Value = $Z_i$ \\
    \> \> \> if $Z_i$ is not partitioned: \\
   \> \> \> \> output Key = $Z_i$, Value = $V_i$ \\
    \> \> \> else: \\
   \> \> \> \> output Key = $Z_i^+$, Value = $V_i$ for a random processor \\
\> 3 Add: \> Changed zones ($Z_i$,$Z_{winner}$) \\
    \> \> \> if $Z_i$ is not partitioned: \\
   \> \> \> \> output Key = $Z_i$, Value = $Z_{winner}$ \\
    \> \> \> else: \\
   \> \> \> \> output Key = $Z_i^+$, Value = $Z_{winner}$ for every processor \\
   \> \> \> \> output Key = $Z_i$, Value = $Z_{winner}$ \\
\> 3 Collate: \> Zone ID as key \\
\> 3 Reduce: \> Emit new zone assignment of each vertex \\
       \> \> \> input Key = $Z_i$ or $Z_i^+$, MultiValue = $V V V V ... Z Z Z ...$ \\
       \> \> \> $Z_{new}$ = min($Z_i$ or $Z_i^+$,Z,Z,Z,...) \\
       \> \> \> partition $Z_{new}$ if number of $V$ > threshhold \\
       \> \> \> for each $V$ in MultiValue: \\
      \> \> \> \> output Key = $V_i$, Value = $Z_{new}$

@@

 \caption{MapReduce algorithm for connected component labelling.}

 \label{fig:cc}
\end{figure}

The algorithm begins (before the iteration loop) by assigning each
vertex to its own component or ``zone'', so that $Z_i$ = $V_i$.  Each
iteration will grow the zones, one layer of neighbors at a time.  As
zones collide due to shared edges, a winner is chosen (the smaller
zone ID), and vertices in the losing zone are reassigned to the
winning zone.  When the iterations complete, each zone will have
become a fully connected component.  The algorithm thus finds all
connected components in the graph simultaneously.  The number of
iterations required depends on the largest diameter of any component
in the graph.

The first {\it map()} operation emits the vertices in each edge as
keys, with the edge as a value.  The current zone assignment of each
vertex is added to the set of key/value pairs.  The first collate()
operation collects all the edges of a vertex and its zone assignment
together in one multi-value.  The first reduce operation then re-emits
each edge, tagged by the zone assignment of one of its vertices.

Since each edge was emitted twice, the second collate operation
collects the zone assignments for its two vertices together.  If the
two zone IDs are different, the second reduce operation chooses a
winner (the min of the 2 IDs), and emits the loser ID as a key, with
the winning ID as a value.  If no zone ID changes are emitted, the
algorithm is finished, and the iteration exits.

The third map() operation inverts the vertex/zone key/value pairs to
become zone/vertex pairs.  The third add operation adds the changing
zone assignments to the set of key/value pairs.  The third collate()
can then collects all the vertices of a zone and zero or more
reassignments for the zone ID.  Since a zone could collide with
multiple other zones on the same iteration due to shared edges, the
new zone ID becomes the minimum ID of any of the neighboring zones.
If no zone reassignment value appears in the multi-value, the zone ID
is unchanged.  The final reduce() a key/value pair for each vertex in
the zone, with the vertex as a key and the new zone ID as a value.

Note that if a graph has only a few components, then the third collate
operation, which keys on the zone ID, may generate a few very large
key/multi-value (KMV) pairs.  For example, if the graph is fully
connected, then on the last iteration, a single KMV pair will contain
all vertices in the graph and be assigned to one processor.  This
imbalance in memory and computational work can lead to poor parallel
performance of the overall algorithm.  To counter this effect, the
various operations of stage 3 include extra logic.  The idea is to
partition zones whose vertex count exceeds a user-defined threshhold
into $P$ sub-zones, where $P$ is the number of processors.  The 64-bit
integer that stores the zone ID also stores a bit flag indicating the
zone has been partitioned and a set of bits that encode the processor
ID.

During the third map() operation, if the zone has been partitioned,
then the vertex is assigned to a random processor and the processor ID
bits are added to the zone ID, as indicated by the $Z_i^+$ notation in
Figure \ref{fig:cc}.  Likewise, if $Z_i$ has been partitioned, the
third add() operation emits the zone ID change key/value pair
($Z_i$,$Z_{winner}$) as ($Z_i^+$,$Z_{winner}$).  In this case
($Z_i$,$Z_{winner}$), is emitted not once, but $P+1$ times, once for
each processor, and once as if $Z_i$ had not been partitioned (for a
reason discussed below).

This additional partitioning logic means that the third collate()
operation, which keys on zone IDs, some of which now include processor
bits, will collect only a $1/P$ subset of the vertices in large zones
onto each processor.  But the multivalue on each processor contains
all the zone-reassignments relevant to the unpartitioned zone.  This
allows the third reduce operation to change the zone ID (if necessary)
in a consistent manner across all $P$ multi-values that contain the
zone's vertices.  When the reduce() operation emits new zone
assignments for each vertex, the zone retains its partitioned status,
and the partition bit is also explicitly set if the vertex count
exceeds the threshhold for the first time.

Note that this logic does not guarantee that the partition bits of the
zone IDs for all the vertices in a single zone will be set
consistently on a given iteration.  For example, an unpartitioned zone
with a small ID may consume a partitioned zone.  The vertices from the
partitioned zone will retain their partitioned status, but the
original vertices in the small zone may not set the partition bit of
their zone IDs.  On subsequent iterations, the third add() operation
emits $P+1$ copies of new zone reassignments for both partitioned and
unpartioned zone IDs, to insure all vertices in the zone will know the
reassignment information.

@@

footnote on NP-complete for maximal; MapReduce is unlikely to
help with that issue.
